---
title: "Creating methods in R"
author: Rakesh Poduval, `r Sys.Date()`
output:
  html_document: 
    toc: FALSE
    toc_depth: 2
    toc_float: TRUE
    theme: yeti
    highlight: monochrome
---

# Specification
Methods in R is a very strong and popular feature. The document provides a simple example with a function we can use to exploratory analysis.

# `explore()` function

In baseR we have the function called `summary()` which is very  power full and works with almost all type of R objects. How does this works ? How come the same function works with vectors, data frame, lm (linear models) object and so on...

Let us understand this with a simple example:

Suppose we need a modified version of the above `summary()` which will provide us with more quantile information regarding and regardless of vectors/dataframe as input this should work.

Let's define the 2 separate function to handle the following cases:

* *`exploreX()`* -> for vectors
* *`exploreData()`* -> for data.frame

## Step 1 - defining the method
define the parent function properly.

```{r}
explore <- function(x, ...) {
  UseMethod("explore")
}
```

## Step 2 - defining the default operation
define the default operation, in this case we expect non numeric vectors.

```{r}
#' explore.default
#' @param x an input vector
#' @param rnd roundind off the results
#'
explore.default <- function(x, rnd = 4, ...) {
  
  # quick check: input must be numeric
  stopifnot(class(x) %in% c('integer', 'numeric'))
  
  # defining the quantiles for the function
  prbs <- c(0, 0.05, 0.25, 0.50, 0.75, 0.95, 1)
  nams <- c("Min", "P05", "P25", "Med", "P75", "P95", "Max")
  qnts <- rbind.data.frame(round(quantile(x, probs = prbs, na.rm = T, names = F), rnd))
  names(qnts) <- nams
  
  # returning output as a data.frame
  cbind.data.frame(
    N = length(x), Mis = sum(is.na(x) | x < 0), 
    Avg = round(mean(x, na.rm = T), rnd), 
    qnts, Std = round(sd(x, na.rm = T), rnd))
}
```

## Step 3 - extending to data.frame
The function is extended to react to a data.frame

```{r}
explore.data.frame <- function(x, rnd = 4, colNames = colnames(x), ...) {
  
  # quick check: input data.frame must have 1 row
  stopifnot(nrow(x) > 0, colNames %in% colnames(x))
  
  # excluding non numeric columns from the analysis
  c <- colnames(x[!sapply(x, class) %in% c('character', 'factor')])
  c <- intersect(c, colNames)
  
  # confirming existence of numeric columns
  stopifnot(length(c) > 0)
  
  # applying the default function to each numeric columns of a data.frame
  if (length(c) == 1) { # special case for a single column
    out <- explore.default(x[, c], rnd = rnd)
    rownames(out) <- c
  } else { # when there are multiple columns to analyse
    out <- do.call(rbind, apply(x[c], 2, explore.default, rnd = rnd))
  }
  return(out)
}
```

## examples
Now the function can be applied to both `vectors` and and a `data.frame`

```{r}
explore(iris)
# explore(iris$Species)
explore(iris$Sepal.Length)
explore(iris, colNames = c('Sepal.Length', 'Petal.Length', 'Species'))
```

Note: With the split function you are able to apply it with different groups
```{r}
# grouped by Species
lapply(split(iris, iris$Species), explore)
```

# Pros and Cons

**pros**

* works for multiple grouping factors with multiple variables.

**cons**

* not that fast
* won't work with `tidyverse` grouping feature.

