---
title: "Introduction to tidyverse"
author: Rakesh Poduval, `r Sys.Date()`
output:
  html_document: 
    toc: FALSE
    toc_depth: 2
    toc_float: TRUE
    theme: yeti
    highlight: monochrome
    number_sections: false
editor_options: 
  chunk_output_type: console
---

```{r setup, echo = F, include = F, cache = F}
library(devtools)
library(tidyverse)
```

# Specification

The tidyverse is an opinionated collection of R packages designed for data science. All packages share an underlying design philosophy, grammar, and data structures. 

Install the complete tidyverse with: `install.packages('tidyverse')` 

For online documentation refer : `browseVignettes(package = "tidyverse")`

Book for reference : [R for Data Science by Hadley Wickham](https://r4ds.had.co.nz)

The book is really good and is recommended strongly. 

# Examples
The package is extremely powerful and easy to work with. Lets take the so called `iris` data for this. Hadley uses `%>%` which he calls _pipes_ to perform `tidyverse` operations. You can use the keyboard shortcut `ctrl + shift + M` for the pipe.

## Data manupulation

Lets start with simple operations like sorting the data.

```{r, message = F}
## order with respect to a column, by default it is in ascending order
iris %>% arrange(Sepal.Length) %>% head(5)
iris %>% arrange(desc(Sepal.Length)) %>% head(5)
```

Renaming existing columns. 

```{r, message = F}
## renaming column name
iris %>% rename(sepalWidth = Sepal.Width) %>% head(3)
```

The select can be used in many ways based on the requirement. Please go through the online reference for more details.

```{r, message = F}
## selecting columns from a data. This also can be used for renaming
iris %>% 
  select(sepalWidth = Sepal.Width, Petal.Width) %>% 
  head(3)
```

Adding a new column can be dine in multiple ways. `mutate()` is useful to add new columns that are functions of existing columns. `transmute()` eliminates all other columns not mentioned in it.

```{r, message = F}
# transform adds a new column
iris %>% 
  transform(ratio = Sepal.Length/Sepal.Width) %>% 
  head(3)

# mutate adds new columns that are functions of existing columns
iris %>%
  mutate(ratio = Sepal.Length/Sepal.Width,
         adjusted = ratio * Petal.Length) %>% 
  head(3)

# transmute eliminates rest of the columns
iris %>% 
  transmute(ratio = Sepal.Length/Sepal.Width) %>% 
  head(3)
```

To subset a data use the `filter()` command

```{r, message = F}
iris %>% 
  filter(Sepal.Length >= 5, Sepal.Width <= 6) %>% 
  head(5)
```

The group by is one of my favourite feature.

```{r, message = F}
# descriptives of single variable
iris %>% 
  group_by(Species) %>% 
  summarise(N = n(), 
            mean = mean(Sepal.Length),
            median = median(Sepal.Length))

# descriptive of all variables
iris %>% 
  group_by(Species) %>% 
  summarise_all(c('mean'))
# iris %>% group_by(Species) %>% summarise_all(c('mean', 'median'))
```

Reshaping is widely used in our day to day operations
```{r }
# long data
iris %>% gather(variable, value, -Species) %>% head(5)

# wide data
  
```

we can even see add graphs at the end of any pipe operation
```{r, results = 'asis', message = F}
iris %>% 
  filter(Sepal.Length > 5) %>% 
  mutate(ratio = Sepal.Length/Sepal.Width) %>% 
  ggplot(aes(Sepal.Length, ratio)) +
  geom_point(aes(color = Species))
```

## Linear modelling

Lets say we want to check the sensitivity of an explanatory variable after fitting a linear model. `modelr` package is very useful for this purpose which is currently under development and not added to the `tidyverse` package. But I expect it to be added in the later versions.
In the below example first 

* we fit a model
* prepare sensitivity data
* predict with the model
* plot sensitivity plot
* check if predictions are significantly different for different Species

```{r }
library(modelr)
  
fit <- lm(Sepal.Length ~ . , data = iris)

data_grid(iris, 
          Species, 
          Petal.Length = seq_range(Petal.Length, n = 5), 
          .model = fit) %>% 
  add_predictions(fit, var = "pred") %>% 
  ggplot(aes_string("Petal.Length", "pred", color = "Species")) +
  geom_line() +
  geom_point()
```

# Conclusion
This helps in narrating the complete story of the operations one is performing with the data and also add some interesting results at the end.
